Classical Simulation of Quantum Error Correction in a Fibonacci Anyon Code

### 2015-06-02 fib2.tex
remove some comments from the latex source
cluster -> component; fix caption

We use a phenomenological model of Fibonacci anyon dynamics which neglects any 
microscopic details of the system. 
This is consistent with the principles of topologically 
ordered systems and anyonic physics, where the key universal features describing the 
anyon model correspond to large length-scale physics, while the microscopic physics plays 
a less important (and non-universal) role.
Our simulation is in effect that of a $(2+1)$-dimensional topological quantum field theory, which allows for topological processes such as pair-creation, braiding, and fusion of anyonic charges. This is modelled on a periodic $L\times L$ square lattice $\Lambda$ of \emph{tiles} on which anyons can reside.  For more details of an analogous model, see~\cite{Brell2013}.
%Effectively, we simulate a $(2+1)$-dimensional topological quantum field theory on a discrete space-time.
%As such, we model the system simply as a 
%periodic $L\times L$ square lattice $\Lambda$ of sites upon which sets of anyons can 
%reside. 
%Allowed anyon dynamics (pair-creation, braiding, and fusion) are considered to act 
%locally on edges of $\Lambda$. 
%For more details of an analogous model, see~\cite{Brell2013}.
\dude{something about observables vs. states.}

@@

Our error-correction algorithm is
based on a hierarchical clustering algorithm~\cite{Hastie2009},
and follows a similar strategy to the hard-decision renormalization group (RG) decoder~\cite{Bravyi2011}. 
The decoder proceeds by measuring the occupations of 
anyons at each site of the lattice, producing a \emph{syndrome} of occupied sites. Following this, it forms clusters of nearby anyons and 
%fusing the anyons within each cluster. 
measures the total charge within each cluster.
Clusters with trivial charge are discarded, and
the rest are joined (agglomeratively~\cite{Hastie2009})
at linearly increasing length scales. This iterative process concludes when there is at most a single cluster remaining (see \Fref{f:decode}).
Note that unlike the case of abelian anyons, the decoder
cannot determine the charge of each cluster given only the syndrome information (as in \cite{Bravyi2011}),
and so will query (measure) the system to find this.
Our simulation proceeds in this way as a dialogue between decoder
and system, terminating when either the decoder itself terminates or when it performs a homologically nontrivial operation, thereby altering the codespace (and failing to perform successful information recovery).
%or the clustering length scale exceeds the size of the lattice.
%\cggb{our simulation may terminate when the decoder performs a nontrivial operation, but here we're just describing the decoder, which can't know that.}
%\dude{in our sim the decoder stops when there is one cluster. Ie., it never gives up.}
%\cggb{A few more details? What is our length schedule?} 
%\cggb{How does our decoder relate to other decoding schemes used in~\cite{Wootton2013,Brell2013,Hutter2014}?} 
Note that, as was found in Ref.~\cite{Brell2013}, we expect that our qualitative results may be reproduced by most alternative families of decoders, although it may not be clear in all cases how to extend these decoders to the non-abelian setting.
The advantage of using a clustering decoder is 
its simplicity and flexibility, and the fact that its clustering scheme is compatible with the structure in the noise processes that allows us to classically simulate them.

\begin{figure}[t!]
\begin{center}
	\includegraphics[width=1.0\columnwidth]{pic-decode.pdf}
\caption{The decoder works by maintaining a set of disjoint clusters as rooted trees.
(a) At the initial clustering stage, these trees are formed from neighboring sites that contain charges. Within each cluster, the charges are transported to the root of the tree (chosen arbitrarily), and their combined charge measured. The direction of transport (towards the root) is denoted by arrows.
(b) At each successive round, all trees are grown in every direction, and overlapping trees are joined. Again, any charges within a cluster are transported to the root of the tree and measured. \cggb{Do we delete the clusters that do not contain any charges (fuse to vacuum) or leave them around?}
}
\label{f:decode}
\end{center}
\end{figure}

%\begin{figure*}[th!]
%\begin{center}
%	{\scriptsize (a)
%	\includegraphics[width=0.45\columnwidth]{pair-create.pdf}
%	\hskip 4pt
%	(b)
%	\includegraphics[width=0.45\columnwidth]{syndrome-1.pdf}
%	\hskip 4pt
%	(c)
%	\includegraphics[width=0.45\columnwidth]{syndrome-2.pdf}
%	\hskip 4pt
%	(d)}
%	\includegraphics[width=0.45\columnwidth]{syndrome-2.pdf}
%\caption{\stf{polish this a bit} The simulation of noise processes proceeds by a) creating pairs of anyons on neighboring tiles represented by open curves b) joining the curves within a given tile c) measuring the total charge on each tile. \stf{Add another graph that shows how we use syndrome info.}}
%\label{f:simex}
%\end{center}
%\end{figure*}

@@

Below the (bond) percolation threshold for (say) a 2D square lattice, we expect random sets of 
bonds to decompose into separate connected components 
%~\footnote{We use the word cluster to refer both to groupings of nearby anyons as formed by the decoding algorithm and to connected components of a graph under percolation\dude{confusing}, since both of these usages are standard in the literature. The intended meaning will remain clear from context.}
of average size $O(\log(n))$ and variance $O(1)$~\cite{Bazant2000}.
Each noise process in our model is associated with a (randomly distributed) edge, and so 
disconnected components correspond to sets of anyons that could not have interacted at any 
point in their history. 
Although normally anyons can interact by braiding without ever coming 
into contact with one another \dude{um um um}, in this case charge conservation 
implies that each distinct component must have trivial total charge, and so all such interactions are trivial.
This allows us to simulate the braiding processes within each component 
separately. In other words: the quantum state in the fusion space of all anyons factorizes into 
a tensor product over components, and we can compute each factor separately. Since each 
component has size only $O(\log(n))$, we can typically simulate these dynamics efficiently 
because the resulting fusion space has dimension $O(\mathrm{poly}(n))$. There are rare 
random processes that violate this reasoning, but processes such as these are suppressed 
exponentially in the lattice size $L$~\cite{Grimmett1989}.

However, random noise processes are not the only dynamics that we need to consider. We 
must also consider the effect of the error-correction routine itself. This acts iteratively to fuse 
particles on increasing length scales. While this kind of fusion would typically merge clusters, 
forcing us to compute dynamics of larger and larger sets of anyons, large clusters are sparsely distributed 
(and thus unlikely to be merged), and in addition at each length scale the total number of 
anyons present is dramatically reduced by fusion, leading to a smaller number of anyons that 
must be simulated.

Of course, if we were to consider noise so strong that 
the clusters percolate over the entire lattice, we would no 
longer expect to be able to efficiently simulate the noise and recovery processes in the system. 
However, recall that logical errors in our system correspond to processes that act over a non-trivial loop of the torus. 
Below the point at which the decoder begins to create such nontrivial loops, our simulation remains efficient,
%\dude{efficiency has nothing to do with it}
while it begins to declare failure whenever non-trivial logical 
operations occur. This allows us to simulate large lattice sizes 
in the successful error-correction (sub-percolated) regime, including up to $L=128$.

@@

%\stf{Add figure about observables versus states.}

Our concrete simulation proceeds by associating a tile with each site of the 
$L\times L$ periodic square lattice.
In order to keep track of the observables of the system 
we maintain a set of disjoint curves, representing the history of anyons on the lattice.
These curves can also be thought of as a top-down view of the fusion tree:
the anyons comprising the fusion tree are here displayed as points along each curve. 
In general, the curves wind around the two 
dimensional tiles in a haphazard way determined by the progress of the simulation.
This allows a dynamically generated basis for the fusion space that easily allows for tensor 
decomposition into factors corresponding 
to disjoint clusters \dude{confusing use of this word}.
For comparison, previous work 
used a single fixed basis for the fusion space (determined by row-major ordering of 
sites)~\cite{Brell2013}.

\begin{figure}[th!]
\begin{center}
	\includegraphics[width=1.0\columnwidth]{pic-syndrome.pdf}
\caption{Four equivalent representations of a state,
along with two syndrome observables marked with solid and dashed lines:
(a) two pairs are joined in (b).
Deform the curve straight to get (c),
then effect some
braid operations on the state so that (d) the two
observables are now represented on the curve. 
%\cggb{I'm not totally clear on what I'm looking at here. Is this the diagram where we want to explain the whole simulation algorithm? Is the idea that we start with a), which is given by the set of noise processes that have happened, then we connect stuff up arbitrarily into b) to make some fusion tree. Then c) is what's happening after linearizing the picture? And d) is c) after some R and F moves so you go into the standard basis for the fusion tree? Are the red lines between c) and d) just denoting leaves with the same identity or do they have some more precise meaning wrt the braid moves that relate c) and d)?}
}
\label{f:syndrome}
\end{center}
\end{figure}

@@

%Each round of simulation proceeds in four steps as follows (an example is shown 
%in \Fref{f:simex}).
%
%\begin{enumerate}[a)]
%\item The noise process is modeled by a sequence of pair-creation events distributed as 
%a Poisson process across each edge. Curves are created crossing the relevant edge 
%and terminating on the neighboring sites.
%\item We first join curves that participate in the same tile in order to, 
%\cggb{maybe a couple more lines of detail on this and the next point? perhaps also the previous paragraph?}
%\item measure the charge on each tile. Any resulting anyon is left somewhere in the tile.
%\item The decoder examines this syndrome and clusters nearby anyons. Each cluster 
%is then measured (fused), and the decoder continues to cluster and fuse on larger 
%scales until there are no more anyons, or a cluster includes a non-trivial loop on the 
%lattice. Since operators with support on such non-trivial clusters will generally perform an 
%undetectable error on the encoded state, in this case we abandon the simulation and 
%declare failure to error correct.
%\end{enumerate}

>>>>

We use a phenomenological model of Fibonacci anyon dynamics which neglects any 
microscopic details of the system. 
This is consistent with the principles of topologically 
ordered systems and anyonic physics, where the key universal features describing the 
anyon model correspond to large length-scale physics, while the microscopic physics plays 
a less important (and non-universal) role.
Our simulation is in effect that of a $(2+1)$-dimensional topological quantum field theory, which allows for topological processes such as pair-creation, braiding, and fusion of anyonic charges. This is modelled on a periodic $L\times L$ square lattice $\Lambda$ of \emph{tiles} on which anyons can reside.  For more details of an analogous model, see~\cite{Brell2013}.
\dude{something about observables vs. states.}

@@

Our error-correction algorithm is
based on a hierarchical clustering algorithm~\cite{Hastie2009},
and follows a similar strategy to the hard-decision renormalization group (RG) decoder~\cite{Bravyi2011}. 
The decoder proceeds by measuring the occupations of 
anyons at each site of the lattice, producing a \emph{syndrome} of occupied sites. Following this, it forms clusters of nearby anyons and 
measures the total charge within each cluster.
Clusters with trivial charge are discarded, and
the rest are joined (agglomeratively~\cite{Hastie2009})
at linearly increasing length scales. This iterative process concludes when there is at most a single cluster remaining (see \Fref{f:decode}).
Note that unlike the case of abelian anyons, the decoder
cannot determine the charge of each cluster given only the syndrome information (as in \cite{Bravyi2011}),
and so will query (measure) the system to find this.
Our simulation proceeds in this way as a dialogue between decoder
and system, terminating when either the decoder itself terminates or when it performs a homologically nontrivial operation, thereby altering the codespace (and failing to perform successful information recovery).
%or the clustering length scale exceeds the size of the lattice.
%\cggb{our simulation may terminate when the decoder performs a nontrivial operation, but here we're just describing the decoder, which can't know that.}
%\dude{in our sim the decoder stops when there is one cluster. Ie., it never gives up.}
%\cggb{A few more details? What is our length schedule?} 
%\cggb{How does our decoder relate to other decoding schemes used in~\cite{Wootton2013,Brell2013,Hutter2014}?} 
Note that, as was found in Ref.~\cite{Brell2013}, we expect that our qualitative results may be reproduced by most alternative families of decoders, although it may not be clear in all cases how to extend these decoders to the non-abelian setting.
The advantage of using a clustering decoder is 
its simplicity and flexibility, and the fact that its clustering scheme is compatible with the structure in the noise processes that allows us to classically simulate them.

\begin{figure}[t!]
\begin{center}
	\includegraphics[width=1.0\columnwidth]{pic-decode.pdf}
\caption{The decoder works by maintaining a set of disjoint clusters as rooted trees.
(a) At the initial clustering stage, these trees are formed 
from neighboring sites that contain charges. Within each cluster, the 
charges are transported to the root of the tree (chosen 
arbitrarily), and their combined charge measured. The direction of transport 
(towards the root) is denoted by arrows.
(b) At each successive round, all trees are grown in 
every direction, and overlapping trees are joined. Again, any charges 
within a cluster are transported to the root of the 
tree and measured. Any clusters with vacuum total charge are deleted.
\label{f:decode}
\end{center}
\end{figure}

@@

Below the (bond) percolation threshold for (say) a 2D square lattice, we expect random sets of 
bonds to decompose into separate connected components 
of average size $O(\log(n))$ and variance $O(1)$~\cite{Bazant2000}.
Each noise process in our model is associated with a (randomly distributed) edge, and so 
disconnected components correspond to sets of anyons that could not have interacted at any 
point in their history. 
Although normally anyons can interact by braiding without ever coming 
into contact with one another \dude{um um um}, in this case charge conservation 
implies that each distinct component must have trivial total charge, and so all such interactions are trivial.
This allows us to simulate the braiding processes within each component 
separately. In other words: the quantum state in the fusion space of all anyons factorizes into 
a tensor product over components, and we can compute each factor separately. Since each 
component has size only $O(\log(n))$, we can typically simulate these dynamics efficiently 
because the resulting fusion space has dimension $O(\mathrm{poly}(n))$. There are rare 
random processes that violate this reasoning, but processes such as these are suppressed 
exponentially in the lattice size $L$~\cite{Grimmett1989}.

However, random noise processes are not the only dynamics that we need to consider. We 
must also consider the effect of the error-correction routine itself. This acts iteratively to fuse 
particles on increasing length scales. While this kind of fusion would typically merge components, 
forcing us to compute dynamics of larger and larger sets of anyons, large components are sparsely distributed 
(and thus unlikely to be merged), and in addition at each length scale the total number of 
anyons present is dramatically reduced by fusion, leading to a smaller number of anyons that 
must be simulated.

Of course, if we were to consider noise so strong that 
the components percolate over the entire lattice, we would no 
longer expect to be able to efficiently simulate the noise and recovery processes in the system. 
However, recall that logical errors in our system correspond to processes that act over a non-trivial loop of the torus. 
Below the point at which the decoder begins to create such nontrivial loops, our simulation remains efficient,
%\dude{efficiency has nothing to do with it}
while it begins to declare failure whenever non-trivial logical 
operations occur. This allows us to simulate large lattice sizes 
in the successful error-correction (sub-percolated) regime, including up to $L=128$.

@@

Our concrete simulation proceeds by associating a tile with each site of the 
$L\times L$ periodic square lattice.
In order to keep track of the observables of the system 
we maintain a set of disjoint curves, representing the history of anyons on the lattice.
These curves can also be thought of as a top-down view of the fusion tree:
the anyons comprising the fusion tree are here displayed as points along each curve. 
In general, the curves wind around the two 
dimensional tiles in a haphazard way determined by the progress of the simulation.
This allows a dynamically generated basis for the fusion space that easily allows for tensor 
decomposition into factors corresponding 
to disjoint components.
For comparison, previous work 
used a single fixed basis for the fusion space (determined by row-major ordering of 
sites)~\cite{Brell2013}.

\begin{figure}[th!]
\begin{center}
	\includegraphics[width=1.0\columnwidth]{pic-syndrome.pdf}
\caption{Four equivalent representations of a state,
along with two syndrome observables marked with solid and dashed lines:
(a) two pairs are joined in (b).
Deform the curve straight to get (c),
then effect some
braid operations on the state so that (d) the two
observables are now represented on the curve. 
}
\label{f:syndrome}
\end{center}
\end{figure}

@@
